Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-10 ltx/data_paper/DataPaper.tex
added convergence plots
edits to M step

\ifpdf
\usepackage[pdftex]{graphicx}
\else
\usepackage{graphicx}
\fi
\usepackage{color}
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\parham}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}

@@

                                                                                                                                                                                                                                                                                                                                                                                                                                                                            Consider first the maximisation of $\mathcal{Q}$-function with respect to $\xi$, we can write

@@

In order to maximise $\mathcal{Q}$-function with respect to $\boldsymbol\theta$ we can write

@@

We comment on the condition under which $\boldsymbol\Xi_4$ is invertible later on. Now maximizing with respect to $\sigma_{\epsilon}^2$ we get

@@

setting above to zero and solving for $\sigma_{\epsilon}^2$ gives

@@

Finally differentiating with respec to $\sigma_d^2$ we get

@@

setting above to zero and solving for $\sigma_d^2$ we have

@@

$\boldsymbol\Xi_2$ is positive definite under the persistent excitation condition of state vectors, which is guaranteed by the disturbance, $\mathbf e$, and $\tilde{\boldsymbol\Sigma}_e^{-1}$ is positive definite by construction (Note that the inverse of a positive definite matrix is also positive definite \cite{Horn2005}), therefore we have \cite{Coope1994}

@@

meaning \eqref{eq:SecondDerivativexi} is negative and $\hat{\xi}$ represents a maximum. Equation \eqref{eq:SecondDerivativetheta} is negative where $\boldsymbol\Xi_4$ is positive definite requiring $\mathbf q(\mathbf x_t)$ to be of  rank $n_{\theta}$ \cite{Golub1996}. The matrix $\boldsymbol\Psi(\mathbf r')$ defined by \eqref{eq:DefPsi} is of rank $n_{\theta}$ as each row of the matrix is a set of equally spaced Gaussian basis functions with equal widths, consequently each row can be considered as a Riesz basis function for the subspaces they generate \cite{Ferrari2004}. (note that multiplying by $\mathbf \Gamma^{-1}$ doesn't change the rank). However, $\mathbf q(\mathbf x_t)$ defined in \eqref{eq:QmatrixForSigmapoints} cannot be computed analytically as the inner product of a Gaussian and a sigmoid cannot be solved analytically, making it difficult to calculate the rank of $\mathbf q(\mathbf x_t)$ . Therefore, we recommend to check the rank of $\mathbf q(\mathbf x_t)$ at each iteration of the E-step.

From \eqref{eq:SecondDerivativenoise} and \eqref{eq:SecondDerivativedisturbance}, observation noise and disturbance variance estimates clearly represent maxima of the $\mathcal{Q}$-function. 
%  Increasing the joint likelihood function increases the likelihood function $p(\mathbf Y;\boldsymbol \Theta)$ \cite{McLachlan1997}. 
 \begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lll}
\hline \hline
Variable &Equation&Order\\ \hline\\
$\Xi_0, \Xi_1, \Xi_2$&\eqref{eq:Xi0Gibson}, \eqref{eq:Xi1Gibson}, \eqref{eq:Xi2Gibson}&$O(Tn_x^2)$\\
$\Xi_3, \Xi_5$&\eqref{eq:Xi3Ap},\eqref{eq:Xi5Ap} &$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)+O(Tn_{\theta}n_r^2)$ \\
$\Xi_4$&\eqref{eq:Xi4Ap}&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_x^3n_{\theta}n_r^2)$\\
$\hat{\xi}$&\eqref{eq:xihat}&$O(n_x^3+n_{\theta}^3)$\\
$\hat{\boldsymbol\theta} $&\eqref{eq:thetahat}&$O(n_{\theta}^3)$\\
$\hat{\sigma}_{\epsilon}^2$&\eqref{eq:sigmaepsilonhat}&$O(Tn_x^2n_y)+O(Tn_xn_y^2)$\\
$\hat{\sigma}_{d}^2$&\eqref{eq:sigmadhat}&$O(n_x^3)+O(n_{\theta}^2)$\\
\hline \hline
\end{tabular}}
 \caption {Computational complexity of EM algorithm} 
\label{table:ComputationalComplexity}
\end{center}
\end {table}
\section{Results}
This section demonstrates the identification of the proposed IDE model using simulated data. The observation noise and disturbance were unknown to the estimation algorithm and no assumption were made on the shape of the kernel.  Data was generated employing state space equation (\ref{eq:StateEq}-\ref{eq:QmatrixForSigmapoints}) using a square grid of $n_x= 9 \times 9$ equally spaced ($2.5~mm$) basis functions with $\sigma_{\phi}^2=1.58~mm^2$, assuming the observed field can be fully described by the decomposed IDE model. 100 realizations of 500 ms of data were generated and the first 100 ms was discarded allowing the model's dynamic to reach steady state. The observation noise was set to $\Sigma_{\epsilon}=0.2\times \mathbf I_{n_y}$ and the field disturbance was set to $\mathcal{GP}(\mathbf 0,0.1 \times\gamma(\mathbf{r}-\mathbf{r}'))$ where the covariance function was chosen as a Gaussian with zero mean and the width $\sigma_{\gamma}^2=1.7~mm^2$ . The distance between adjacent sensors was 1.5 mm resulting into $n_y = 196$ observations. The width of the sensor kernels were set to $\sigma_m^2=0.81~mm^2$. The sampling period and the synaptic time constant were set to $T_s = 1~ms$  and $\tau = 10~ms$, respectively. The sigmoidal activation parameters were $\varsigma=0.56~mV^{-1}$ and $v_0=1.8~mV$.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelBasis1.pdf}
\end{center}
\caption{{\bf Connectivity kernel support approximation.} The approximate support of the kernel is shown by the red circle. Black area shows the non-zero values of the support after thresholding. A $5\times5$ grid of basis functions (blue circles) used to reconstruct the connectivity kernel.}
\label{fig:KernelWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/DisturbanceWidthEstimation.pdf}
\end{center}
\caption{{\bf Disturbance support approximation}. }
\label{fig:DisturbanceWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimate.pdf}
\end{center}
\caption{{\bf Connectivity kernel estimates.} (A) Actual kernel. (B) The mean kernel estimate over 100 realisations.}
\label{fig:KernelEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimateCrossSection.pdf}
\end{center}
\caption{{\bf Cross-sections of connectivity kernel estimates.} Actual kernel and the mean kernel estimates are shown with blue and green lines respectively. Each of the estimate from Monte Carlo simulation is shown with black line. The 95\% confidence intervals are shown by shaded red area.}
\label{fig:KernelEstimateCrossSection}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Histogram.pdf}
\end{center}
\caption{{\bf Histograms of the parameter estimates over 100 realizations}. The actual and mean of the estimated parameter values are shown by solid and dotted lines respectively. (A) The histogram of the synaptic dynamic parameter, $\xi$. (B) The histogram of the observation noise variance, $\sigma_{\epsilon}$. (C) The histogram of the disturbance noise variance, $\sigma_d^2$.}
\label{fig:Histogram}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=1]{./Figures/FieldEstimates.pdf}
\end{center}
\caption{{\bf Neural field estimates at different time instants}. In each subplot, true (gray line) and reconstructed neural field (red dots) are plotted versus spatial index.}
\label{fig:FieldEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Q.pdf}
\end{center}
\caption{{\bf Convergence of the EM algorithm.} The mean of the lower bound of the log-likelihood function over 100 realisations. The change in lower bound drops below 0.009 percent after 15 iterations}
%  $Q(\boldsymbol\Theta',\boldsymbol\Theta')=-(T-1)\left[n_y(1+\ln \hat{\sigma}^2_{\epsilon})+ n_x(1+\ln \hat{\sigma}^2_d)\right]$
\label{fig:LowerBound}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

>>>>

\ifpdf
\usepackage[pdftex]{graphicx}
\else
\usepackage{graphicx}
\fi
\usepackage[font=footnotesize]{subfig}
\usepackage{color}
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\parham}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}

@@

                                                                                                                                                                                                                                                                                                                                                                                                                                                                            Consider first the maximisation of $\mathcal{Q}$-function with respect to $\xi$, differentiating equation \eqref{eq:VoldermontwitXivariables} with respec to the synaptic parameter gives

@@

In order to maximise $\mathcal{Q}$-function with respect to the connectivity kernel parameters, $\boldsymbol\theta$, differentiation is taken as follows

@@

The condition under which $\boldsymbol\Xi_4$ is invertible is discussed later in this section. Now maximising the lower bound with respect to $\sigma_{\epsilon}^2$ gives

@@

and setting the above equation to zero and solving for $\sigma_{\epsilon}^2$ gives the result

@@

Finally differentiating the lower bound with respec to $\sigma_d^2$ gives

@@

setting above to zero and solving for $\sigma_d^2$ completes the maximisation step

@@

$\boldsymbol\Xi_2$ is positive definite under the persistent excitation condition of state vectors, which is guaranteed by the disturbance, $\mathbf e$, and $\tilde{\boldsymbol\Sigma}_e^{-1}$ is positive definite by construction (Note that the inverse of a positive definite matrix is also positive definite \cite{Horn2005}), therefore by \cite{Coope1994}

@@

meaning \eqref{eq:SecondDerivativexi} is negative and $\hat{\xi}$ represents a maximum. Equation \eqref{eq:SecondDerivativetheta} is negative where $\boldsymbol\Xi_4$ is positive definite requiring $\mathbf q(\mathbf x_t)$ to be of  rank $n_{\theta}$ \cite{Golub1996}. The matrix $\boldsymbol\Psi(\mathbf r')$ defined by \eqref{eq:DefPsi} is of rank $n_{\theta}$ as each row of the matrix is a set of equally spaced Gaussian basis functions with equal widths, consequently each row can be considered as a Riesz basis function for the subspaces they generate \cite{Ferrari2004} (note that multiplying by $\mathbf \Gamma^{-1}$ has no effect on the rank). However, $\mathbf q(\mathbf x_t)$ defined in \eqref{eq:QmatrixForSigmapoints} cannot be computed analytically as the inner product of a Gaussian and a sigmoid function cannot be solved analytically, making it difficult to calculate the rank of $\mathbf q(\mathbf x_t)$ . Therefore, the rank of $\mathbf q(\mathbf x_t)$ at each iteration of the algorithm should be examined. From \eqref{eq:SecondDerivativenoise} and \eqref{eq:SecondDerivativedisturbance}, observation noise and disturbance variance estimates clearly represent maxima of the $\mathcal{Q}$-function. 
%  Increasing the joint likelihood function increases the likelihood function $p(\mathbf Y;\boldsymbol \Theta)$ \cite{McLachlan1997}. 
 \begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lll}
\hline \hline
Variable &Equation&Order\\ \hline\\
$\Xi_0, \Xi_1, \Xi_2$&\eqref{eq:Xi0Gibson}, \eqref{eq:Xi1Gibson}, \eqref{eq:Xi2Gibson}&$O(Tn_x^2)$\\
$\Xi_3, \Xi_5$&\eqref{eq:Xi3Ap},\eqref{eq:Xi5Ap} &$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)+O(Tn_{\theta}n_r^2)$ \\
$\Xi_4$&\eqref{eq:Xi4Ap}&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_x^3n_{\theta}n_r^2)$\\
$\hat{\xi}$&\eqref{eq:xihat}&$O(n_x^3+n_{\theta}^3)$\\
$\hat{\boldsymbol\theta} $&\eqref{eq:thetahat}&$O(n_{\theta}^3)$\\
$\hat{\sigma}_{\epsilon}^2$&\eqref{eq:sigmaepsilonhat}&$O(Tn_x^2n_y)+O(Tn_xn_y^2)$\\
$\hat{\sigma}_{d}^2$&\eqref{eq:sigmadhat}&$O(n_x^3)+O(n_{\theta}^2)$\\
\hline \hline
\end{tabular}}
 \caption {Computational complexity of EM algorithm} 
\label{table:ComputationalComplexity}
\end{center}
\end {table}
\section{Results}
This section demonstrates the identification of the proposed IDE model using simulated data. The observation noise and disturbance were unknown to the estimation algorithm and no assumption were made on the shape of the kernel.  Data was generated employing state space equation (\ref{eq:StateEq}-\ref{eq:QmatrixForSigmapoints}) using a square grid of $n_x= 9 \times 9$ equally spaced ($2.5~mm$) basis functions with $\sigma_{\phi}^2=1.58~mm^2$, assuming the observed field can be fully described by the decomposed IDE model. 100 realizations of 500 ms of data were generated and the first 100 ms was discarded allowing the model's dynamic to reach steady state. The observation noise was set to $\Sigma_{\epsilon}=0.2\times \mathbf I_{n_y}$ and the field disturbance was set to $\mathcal{GP}(\mathbf 0,0.1 \times\gamma(\mathbf{r}-\mathbf{r}'))$ where the covariance function was chosen as a Gaussian with zero mean and the width $\sigma_{\gamma}^2=1.7~mm^2$ . The distance between adjacent sensors was 1.5 mm resulting into $n_y = 196$ observations. The width of the sensor kernels were set to $\sigma_m^2=0.81~mm^2$. The sampling period and the synaptic time constant were set to $T_s = 1~ms$  and $\tau = 10~ms$, respectively. The sigmoidal activation parameters were $\varsigma=0.56~mV^{-1}$ and $v_0=1.8~mV$.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelBasis1.pdf}
\end{center}
\caption{{\bf Connectivity kernel support approximation.} The approximate support of the kernel is shown by the red circle. Black area shows the non-zero values of the support after thresholding. A $5\times5$ grid of basis functions (blue circles) used to reconstruct the connectivity kernel.}
\label{fig:KernelWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/DisturbanceWidthEstimation.pdf}
\end{center}
\caption{{\bf Disturbance support approximation}. }
\label{fig:DisturbanceWidthEstimation}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimate.pdf}
\end{center}
\caption{{\bf Connectivity kernel estimates.} (A) Actual kernel. (B) The mean kernel estimate over 100 realisations.}
\label{fig:KernelEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/KernelEstimateCrossSection.pdf}
\end{center}
\caption{{\bf Cross-sections of connectivity kernel estimates.} Actual kernel and the mean kernel estimates are shown with blue and green lines respectively. Each of the estimate from Monte Carlo simulation is shown with black line. The 95\% confidence intervals are shown by shaded red area.}
\label{fig:KernelEstimateCrossSection}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Histogram.pdf}
\end{center}
\caption{{\bf Histograms of the parameter estimates over 100 realizations}. The actual and mean of the estimated parameter values are shown by solid and dotted lines respectively. (A) The histogram of the synaptic dynamic parameter, $\xi$. (B) The histogram of the observation noise variance, $\sigma_{\epsilon}$. (C) The histogram of the disturbance noise variance, $\sigma_d^2$.}
\label{fig:Histogram}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=1]{./Figures/FieldEstimates.pdf}
\end{center}
\caption{{\bf Neural field estimates at different time instants}. In each subplot, true (gray line) and reconstructed neural field (red dots) are plotted versus spatial index.}
\label{fig:FieldEstimates}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/Q.pdf}
\end{center}
\caption{{\bf Convergence of the EM algorithm.} The mean of the lower bound of the log-likelihood function over 100 realisations. The change in lower bound drops below 0.009 percent after 15 iterations}
%  $Q(\boldsymbol\Theta',\boldsymbol\Theta')=-(T-1)\left[n_y(1+\ln \hat{\sigma}^2_{\epsilon})+ n_x(1+\ln \hat{\sigma}^2_d)\right]$
\label{fig:LowerBound}
\end{figure}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\centering
% \subfloat[][]{\includegraphics[scale=1]{./Graph/EM/Q.pdf}}\\
\subfloat[][]{\includegraphics[scale=1]{./Figures/ParameterConvergence.pdf}}\\
\centering
\subfloat[][]{\includegraphics[scale=1]{./Figures/KernelConvergence.pdf}}
\caption{{\bf Convergence of parameters}. All parameters converge after 9 iterations. (\textbf a) The logarithm of the mean of synaptic parameter, observation noise variance and field disturbance variance over 100 realisations. (\textbf b) The mean kernel estimates at each iteration of the EM algorithm.}
\label{fig:ParameterConvergence}
\end{figure}
